|

### 2023-01-13 text/ManuscriptDraft.Rmd
Resubmission -first commit

```{r eval = FALSE, echo=FALSE}
library(knitr)
knit('ManuscriptDraft.Rmd')
rmarkdown::render("ManuscriptDraft.Rmd", output_file = "ManuscriptDraft.docx")
system2("open","ManuscriptDraft.docx")
```

@@

A fundamental challenge in pollination ecology is to understand how successful pollination is affected by the timing of flowering, the timing and frequency of pollinator visits, and the abundance and diversity of visitors. Since plant-pollinator interactions concern individual flowers, deciphering the consequences of e.g. shifts in the frequency of flower visits for plant reproductive success should ideally be based on observations of individual flowers throughout their entire flowering period. Due to the overwhelming logistical challenge of collecting such data by direct observation, it has rarely been done. Here, we show that image-based monitoring of field plots at very high temporal resolution can return information on flowering phenology at the level of individual flowers. Further, we present a framework for automatically tracking flowers in time-lapse image series, evaluating tracking accuracy, and identifying unreliable tracks. Our results on individual flower phenology suggest that traditional weekly observations of flower plots are insufficient to capture details in flowering dynamics. We show that the results of our automated method are reliable by comparing with manual tracking of individual flower phenology. Our method could facilitate fine-grained studies on flower phenology and could be an important tool for estimating the sensitivity of plant-pollinator interactions to climate change.

@@

The Arctic is characterized by short growing seasons and for the individual plant, timing of flowering is of critical importance. Flowering too early increases the risk of frost damage to the flowers, while flowering too late limits the time for seed maturation and dispersal before the end of the growing season [@pardee2019; @ELZINGA2007; @inouye2008; @iler2021]. Further, for insect-pollinated plants, flowering must be synchronous with pollinator activity for successful reproduction. Both flower and insect phenology is partly driven by abiotic cues such as snowmelt and temperature and can thus be sensitive to climate change. Changes in flower phenology as a consequence of climate change may indirectly affect the abundance of flower-dependent insect species [@hoye2013] and these effects can be stronger than the direct effects of climate change [@ogilvie2017]. However, the relation between flower phenology and reproductive success is not well understood. For example, @bolmgren2015 found that for an insect-pollinated plant species, early flowering was associated with increased fruit-set but late flowering was associated with higher seed mass. If we are to understand the intricate dynamics of plant-pollinator interactions, tools that aid efficient and comprehensive data collection are essential.

Monitoring flower phenology at high temporal resolution is time-consuming, particularly in logistically challenging environments such as the Arctic. Consequently, flowering phenology of a population is often characterized simply by the date for onset of flowering, typically derived from weekly observations in permanent plots [@Prevey2021]. Such crude estimates may fail to reveal important dynamics in flowering phenology through the season. For example, different phenological factors, such as onset, peak, and end of flowering may respond differently to climate change [@cook2012] and estimation of the degree of mismatch between species may thus be sensitive to the choice of metric [@inouye2019]. The optimal metric for phenology depends on the research question, but often the choice of recording first flowering dates to represent flowering phenology is pragmatic rather than ideal. In fact, first flowering dates may be particularly ill suited for estimating sensitivity to climate change [@miller2008] since they represent the very extreme of the phenological distribution of a population and may be confounded with e.g. the size of the population considered. An accurate estimate of the phenological distribution of a population over the full season would allow for, for example, accurate quantification of resource availability through the season and temporal mismatches between species [@ramakers2020]. However, even detailed information on phenology of a population is still an aggregation of the phenology of the individual members of the population for which important variation may be masked [@inouye2019]. With the full distribution of flowering phenology at the level of individuals, it is possible to extract relevant characteristics of the distribution and test their individual sensitivities towards environmental factors, but acquiring such data requires efficient and preferably automated methods.

Cameras have become an ubiquitous tool for non-invasive and efficient monitoring of natural processes [@howe2017; @glover2019], but the implementation of image-based data collection has outpaced the development of methods for processing and analysing the big data output [@weinstein2018; @farley2018]. Gradually, computer vision and deep learning tools are being developed and applied for automatic processing and extraction of relevant data from images [@willi2019; @tabak2019; @hoye2021]. Indeed, automatic image-based monitoring of flowering phenology coupled with deep learning has facilitated collection of phenology data for specific species at very high temporal resolution [@mann2022]. Yet, phenological responses at the individual level may be indiscernible regardless of the temporal resolution of the data at population level. For example, a shortening of individual flower longevity may not be directly obvious at the population level. In fact, many research questions can only be directly explored on the basis of phenology data at the level of individuals. For example, investigating the association between reproductive success and timing of flowering and/or flower longevity requires such data. Individual level data would also aid the understanding of whether flower visitation rates and/or seed set depends on the timing of flowering for the individual flower. Thus, automated tracking of flowers in image-series of permanent plots to derive individual flower phenology would be ideal.

Most multiple object tracking literature focus on tracking of humans [@luo2021; @milan2016], and these methods not necessarily directly transferable to other use-cases. Many tracking algorithms, such as the Simple online and real-time tracking (SORT) [@sort], rely on Kalmann filtering to predict the future location of an object by estimating its velocity from previous frames [@kalmann]. However, Kalman filtering assumes linear motion of the object and does not handle abrupt motion well. Flowers are constrained in movement by their stalk and violates this assumption. Other tracking approaches, such as DeepSORT [@deepsort2017], use tracking-by-detection in which convolutional neural networks (CNNs) are applied to continuously distinguish individuals through time, thus aiding the tracking algorithm. Flowers, however, are unlikely to be clearly distinguishable and one flower may look less alike between two frames than two flowers in a single frame, which prevents this approach. Finally, some tracking methods require manual initiation of tracks for a fixed number of objects in the first frame and the algorithm then tracks these objects in the subsequent frames [@hu2012]. These methods do not allow objects to enter or disappear from frame during tracking. In full-season monitoring data, flowers will disappear from the frame when they wilt, in which case their track should be terminated. Oppositely, new flowers should continuously be tracked as they bloom throughout the season. Additionally, a flower tracking method must handle flowers disappearing and appearing at the edges of the frame. Clearly, individual flower tracking is a challenging task.

Here, we present and evaluate a simple multiple object tracking algorithm for automatic tracking of individual flowers in image-series. Our flower tracking algorithm assigns individual flowers to tracks based on the minimum distance between centroids of bounding boxes and does not rely on prediction of future positions. Further, the method allows both initiation and termination of tracks during tracking. We describe three user-defined parameters to help optimise tracking of flowers in particular. We use time-lapse image-series from full-season monitoring of two widespread Arctic flowering plant species *Dryas integrifolia* and *D. octopetala*. For complex scenes with many flowers in close vicinity to each other there is a substantial risk of tracking errors regardless of the tracking method being used. In such cases, it may still be possible to reliably track a subset of the flowers, which are spatially isolated in the frames. For this, we suggest a conservative filtering approach based on the distance between tracks within the frame. Overall, we show that information on phenology at the level of individual flowers can be automatically derived from image-based monitoring of flower plots.

@@

## Study site and species

## The image series and flower annotations

Images were collected throughout the full growing seasons. While the original time-lapse frequency was higher, the basis for this study was 1-hour interval series. We limited these series to include only the flowering season and, to limit the workload for manual annotation, randomly sampled images from each series. The number of images for each series after sampling is given in Table 1. All flowers in bloom within each of these image were manually annotated using the rectangular bounding box tool in the VIA VGG annotation software [@dutta2019] and each individual flower was assigned a unique ID that was maintained throughout its lifetime. These annotations constitute our ground truth tracks. The series varied substantially in the number of flowers they contained (Table 1). Examples of annotated images from the 2019 NARS C series at three different times in the flowering season are given in Figure 1. To demonstrate how individual flower tracking can result in detailed representations of flower phenology across the season, we use the ground truth annotations to show the variation in duration of flowering for each series.

**Table 1:** Number of images and number of individual flowers for each of the five time-lapse image series.

![](../figures/CameraOverview.png){width="40%"}

![](../figures/AnnotationExamples_v2_4.png){width="20%"}

**Figure 1:** Examples of annotated images for 2019 NARS C. A: Onset of flowering at DOY 145. B: Peak flowering at day of year (DOY) 151 (52 annotated flowers). C: 6 days after peak flowering (DOY 157). Flower bounding boxes are colour-coded by flower ID. D: Centroids of all flower annotations in the series, coloured by flower ID, showing how flowers may overlap in space.

## Automatic flower tracking

We built a framework for tracking, filtering, and evaluating tracking of objects in time-lapse image series. The tracking algorithm (suppl. 1) tracks objects based on distances between centroids of bounding boxes, that is, objects with minimum displacement between frames are associated. The algorithm has a set of user-adjusted parameters that can optimize tracking accuracy. The parameters are particularly relevant for tracking of objects that are constrained to a specific area such as flowers. However, the tracking algorithm could be used to track any objects. The speed of the tracking algorithm depends on the computational power available as well as the number of objects that are being tracked, but the time for computation will be negligible in most use-cases. Tracking of a series containing 85 objects ran at \~0.02 seconds per frame on a standard laptop. This, together with the fact that the algorithm handles frames sequentially, means that it can be applied both offline (on a set of detections/annotations that have already been produced) or online (real-time tracking frame per frame).

Several issues complicate the task of tracking individual flowers through time-lapse image series. First, as flowers move, they may temporarily occlude the view of other flowers and flowers close to the edge of the frame may move in and out of view between images. This can cause a track to be lost and a new track to be erroneously initiated when the flower reappears (Fig. 2A), a tracking error known as fragmentation. However, if a flower reappears in the same area as a flower is already being tracked after disappearing in only a few frames, it is a reasonable assumption that it is the same individual and not that the old flower wilted/disappeared and a new one developed. The parameter **max gap** sets the number of frames a flower can be absent before reappearing without the track ending and a new initiating. If a given flower has not been detected in a number of frames exceeding the max gap size, the track will be terminated. Concurrently, this can deal with potential false negatives, as a track that is missing a bounding box will not be lost.

Second, a standard centroid tracking algorithm will associate a point with a track based on minimum distance only, disregarding the absolute distance between centroids, which may lead to tracking errors (Fig. 2B). However, as flowers are constrained in their displacement, we can assume a maximum distance between points and force the initiation of tracks for points that exceed this threshold. We set this threshold with the parameter **max distance,** measured as the euclidean distance in number of pixels between objects in neighbouring frames.

Third, reliable tracking based on distance requires that movement of individuals are captured at sufficient frequency, such that each tracks are not switched between frames. Stand-alone image-based monitoring of the relatively slow process of flower development through full growing seasons (several months) is done with low frame rate, limited by power consumption and storage capacity both on and off the camera. Here, we used several images per day and while this temporal resolution captures the phenology of the flowers at very high detail, they fail to capture the full movement of flowers. For example, as the wind shifts, the flower heads change direction, which in the image series happens instantaneously (i.e., between two consecutive frames) (Fig. 2C). The fact that the frame rate is insufficient to capture smooth movement means that bounding boxes around a single individual in two consecutive frames cannot be associated based on overlap (e.g. intersection over union) as there often is no overlap. With a high number of flowers in close vicinity to each other, there's a substantial risk of identity switching between flowers as they move among each other (Fig. 1B). Additionally, as the flowers are constrained by their stalk, there is a limit to the distance they can move, which prevents prediction of future location of an individual based on its previous trajectory. Here, we introduce the tracking parameter **running mean**, with which we can base tracking on the distance between a point in the current frame and the running mean of the positions of a number of previous points in a track, taking advantage of the fact that a flower moves around a central point. Increasing the value for this parameter decreases the weighting of movement between individual frames. Setting the parameter to 1 means that tracking is based on the distance to the centroid in the previous frame only.

![Figure 2: Centroid tracking may produce erroneous associations when objects move or disappear between frames. A: A flower temporarily disappears from view. If gaps in tracks are not allowed during tracking, a new track will erroneously be initiated when the flower reappears. B: A flower disappears from view and another reappears between two frames. If no threshold for maximum distance between an object and a track is set, the reappearing flower will be linked to the track of the disappearing flower based on shortest distance despite the absolute distance being large. C: A flower is linked to the track of another flower. As flowers move around a central point, it is relevant to base tracking on a running mean of previous positions. Dots show position of flower in previous frame. Crosses show running mean of positions in x \> 1 number of previous frames. Dots and crosses coloured according to track identity.](../figures/ErrorFigure2.png){width="100%"}

### Input data for the tracking algorithm

### Identifying optimal user parameters

To explore the effect of the three tracking parameters and to identify the optimal combination of parameters for our use-case of tracking flowers, we applied the tracking algorithm on each of the five image series with every combination of a range of values for each parameter (3179 combinations): 0-160 with a step size of 10 for max gap; 1-160 with a step size of 10 for running mean; 0-1000 with a step size of 100 for max distance. Note that max distance set to zero ignores the parameter altogether. We identified the setting(s) that returned the most accurate tracks for each series and compare the tracking results between optimal settings and all parameters ignored (imax gap = 0, running mean = 1, max distance = 0).

## Evaluating tracking performance

@@

where fn, fp, and mm are the number of false negatives, false positives, and track mismatches, respectively, for frame t divided by the total number of ground truth annotations across all frames. The MOTA score ranges from -Inf to 1 with 1 being perfect detection and tracking. As tracking accuracy is commonly evaluated between manual annotations and detections, produced, for example, by a convolutional neural network (CNN), the MOTA score included false positive and false negatives as errors, but in our case, since we are tracking on ground truth data and our annotations and detections therefore are practically identical, these variables can be ignored.

@@

## Identifying the most reliably tracked flowers

In order to be able to identify the most reliably tracked flowers without the need for manual annotation, we use a conservative filtering method that extracts the most trustworthy tracks from a scene based on spatial distance between tracks. We derive the distance between tracks based on the centroids of points included in a track produced by the tracking algorithm. When tracks consist of only a single point, the coordinates of the point is used as the centroid. For tracks consisting of two points. we use the midpoint of the line between them. For tracks consisting of three points or more, we calculate the convex hull of all the points included in the track and use the centroid of this polygon. For the special case where a track contains more than two points but where the points are collinear, we use the midpoint of the line through the points. We hereafter refer to these centroids as track centroids.

@@

Manual tracking of individual flowers derived from image-based plot monitoring enabled detailed information on flower phenology at the level of individuals with substantial variation in flowering duration between flowers (Fig. 3). Further, our tracking algorithm was able to track individual flowers with consistently high accuracy (MOTA $\geq$ 0.981) across all series. In all cases, all flowers were tracked with more than 80% accuracy (Table 2). In fact, two series were tracked without any errors (2018 THUL B and 2018 NARS D). Tracking with the three tracking parameters ignored (max gap = 0, max distance = 0, running mean = 1) returned high MOTA scores across series, but for the three series that were not tracked perfectly, utilizing the tracking parameters increased the MOTA score in all cases.

Importantly, the tracking that returned the lowest number of mismatches did not necessarily result in a number of tracks equal to the number of flowers in a plot. For example, for the series with the most flowers, 2019 NARS C, one combination of parameters (max gap = 20, max distance = 600, running mean = 1), returned the lowest number of mismatches (31), but the number of tracks were lower than the actual number of flowers (71 vs. 85). One combination of parameters did return a number of tracks equal to number of flowers (max gap = 10, max distance 400, running mean = 10), but also increased the number of mismatches to 91. For 2019 NARS D, several combinations of parameters returned a perfect MOTA score, but still some combinations returned a number of tracks lower than the number of flowers. This demonstrates the significance of a special case of tracking error, namely merging of tracks, which is not accounted for in the MOTA score. In Figure 4, we exemplify the two types of errors, identity shifts and merges.

>>>>

```{r eval = FALSE, echo=FALSE}
library(knitr)
knit('ManuscriptDraft.Rmd')
rmarkdown::render("ManuscriptDraft.Rmd", output_file = "ManuscriptDraft_Resubmission.docx")
system2("open","ManuscriptDraft_Resubmission.docx")
```

@@

 1. A fundamental challenge in pollination ecology is to understand how successful pollination is affected by the timing of flowering, the timing and frequency of pollinator visits, and the abundance and diversity of visitors. Since plant-pollinator interactions concern individual flowers, deciphering the consequences of e.g. shifts in the frequency of flower visits for plant reproductive success should ideally be based on observations of individual flowers throughout their entire flowering period. Due to the overwhelming logistical challenge of collecting such data by direct observation, it has rarely been done.

 2. Here, we show that image-based monitoring of field plots at very high temporal resolution can return information on flowering phenology at the level of individual flowers. Further, we present a framework for automatically tracking flowers in time-lapse image series, evaluating tracking accuracy, and identifying unreliable tracks.

 3. Our results on individual flower phenology suggest that traditional weekly observations of flower plots are insufficient to capture details in flowering dynamics. We show that the results of our automated method are reliable by comparing with manual tracking of individual floral phenology.

 4. Our method could facilitate fine-grained studies on flower phenology and could be an important tool for estimating the sensitivity of plant-pollinator interactions to climate change.

@@

The Arctic is characterized by short growing seasons and for the individual plant, timing of flowering is of critical importance. Flowering too early increases the risk of frost damage to the flowers, while flowering too late limits the time for seed maturation and dispersal before the end of the growing season [@pardee2019; @ELZINGA2007; @inouye2008; @iler2021]. Further, for insect-pollinated plants, flowering must be synchronous with pollinator activity for successful reproduction. Both flower and insect phenology is partly driven by abiotic cues such as snowmelt and temperature and can thus be sensitive to climate change. Changes in floral phenology as a consequence of climate change may indirectly affect the abundance of flower-dependent insect species [@hoye2013] and these effects can be stronger than the direct effects of climate change [@ogilvie2017]. However, the relation between floral phenology and reproductive success is not well understood. For example, @bolmgren2015 found that for an insect-pollinated plant species, early flowering was associated with increased fruit-set but late flowering was associated with higher seed mass. If we are to understand the intricate dynamics of plant-pollinator interactions, tools that aid efficient and comprehensive data collection are essential.

Monitoring floral phenology at high temporal resolution is time-consuming, particularly in logistically challenging environments such as the Arctic. Consequently, flowering phenology of a population is often characterized simply by the date for onset of flowering, typically derived from weekly observations in permanent plots [@Prevey2021]. Such crude estimates may fail to reveal important dynamics in flowering phenology through the season. For example, different phenological factors, such as onset, peak, and end of flowering may respond differently to climate change [@cook2012] and estimation of the degree of mismatch between species may thus be sensitive to the choice of metric [@inouye2019]. The optimal metric for phenology depends on the research question, but often the choice of recording first flowering dates to represent flowering phenology is pragmatic rather than ideal. In fact, first flowering dates may be particularly ill suited for estimating sensitivity to climate change [@miller2008] since they represent the very extreme of the phenological distribution of a population and may be confounded with e.g. the size of the population considered. An accurate estimate of the phenological distribution of a population over the full season would allow for, for example, accurate quantification of resource availability through the season and temporal mismatches between species [@ramakers2020]. However, even detailed information on phenology of a population is still an aggregation of the phenology of the individual members of the population for which important variation may be masked [@inouye2019]. With the full distribution of flowering phenology at the level of individual flowers, it is possible to extract relevant characteristics of the distribution and test their individual sensitivities towards environmental factors, but acquiring such data requires efficient and preferably automated methods.

Cameras have become an ubiquitous tool for non-invasive and efficient monitoring of natural processes [@howe2017; @glover2019], but the implementation of image-based data collection has outpaced the development of methods for processing and analysing the big data output [@weinstein2018; @farley2018]. Gradually, computer vision and deep learning tools are being developed and applied for automatic processing and extraction of relevant data from images [@willi2019; @tabak2019; @hoye2021]. Indeed, automatic image-based monitoring of flowering phenology coupled with deep learning has facilitated collection of phenology data for specific species at very high temporal resolution [@mann2022]. While automatic detection and counting of flowers in image-series, as demonstrated in @mann2022, can produce detailed information about floral biology, phenological responses at the level of individual flowers may be indiscernible regardless of the temporal resolution of the data at population level. For example, a shortening of individual flower longevity may not be directly obvious at the population level. In fact, many research questions can only be directly explored on the basis of phenology data at the level of individual flowers and such information is critical for investigating the association between reproductive success and timing of flowering and/or flower longevity. Individual level data would also aid the understanding of whether flower visitation rates and/or seed set depends on the timing of flowering for the individual flower. Thus, automated recording of floral phenology by tracking the presence of individual flowers in image-series of permanent plots would be ideal and the feasibility of this as a second step subsequent of flower detection was explored in this study.

![Figure 1: Tracking individual flowers through time. A: A time-lapse sequence of five frames showing presence of five individual flowers. Flower a (orange) is present in all frames, while the other flowers are only present part of the sequence and/or have gaps in their presence (e.g. because of occlusion or because they have temporarily disappeared from frame of view). Correct tracking of individual flowers results in information about the distribution of each individual flower in space (B) and time (C).](../figures/flowertrackingOverview.png){width="30%"}

Most multiple object tracking literature focus on tracking of humans [@luo2021; @milan2016], and these methods not necessarily directly transferable to other use-cases. Many tracking algorithms, such as the Simple online and real-time tracking (SORT) [@sort], rely on Kalmann filtering to predict the future location of an object by estimating its velocity from previous frames [@kalmann]. However, Kalman filtering assumes linear motion of the object and does not handle abrupt motion well. Flowers are constrained in movement by their stalk and violate this assumption. Other tracking approaches, such as DeepSORT [@deepsort2017], use tracking-by-detection in which convolutional neural networks (CNNs) are applied to continuously distinguish individuals through time, thus aiding the tracking algorithm. Individual flowers, however, are unlikely to be clearly distinguishable and one flower may look less alike between two frames than two flowers in a single frame, which prevents this approach. Finally, some tracking methods require manual initiation of tracks for a fixed number of objects in the first frame and the algorithm then tracks these objects in the subsequent frames [@hu2012]. These methods do not allow objects to enter or disappear from frame during tracking. In full-season monitoring data, flowers will disappear from the frame when they wilt, in which case their track should be terminated. Oppositely, new flowers should continuously be tracked as they bloom throughout the season. Additionally, a flower tracking method must handle flowers disappearing and appearing at the edges of the frame. Clearly, individual flower tracking is a challenging task.

Here, we present and evaluate a simple multiple object tracking algorithm for automatic tracking of individual flowers in image-series. Our tracking algorithm is based on centroid tracking and assigns individual objects to tracks based on the minimum distance between centroids of bounding boxes and does not rely on prediction of future positions. Further, the method allows both initiation and termination of tracks during tracking. We describe three user-defined parameters incorporated into the algorithm, demonstrate their effect on simulated data, and discuss why they may help optimise tracking of flowers in particular. We apply the algorithm on annotations of real flowers in time-lapse image-series from full-season monitoring of two widespread Arctic flowering plant species *Dryas integrifolia* and *D. octopetala*. For complex scenes with many flowers in close vicinity to each other there is an increased risk of tracking errors regardless of the tracking method being used. For such cases, we suggest a conservative filtering approach based on the distance between flower tracks within the frame in order to identify tracks that are spatially isolated and likely to be tracked with certainty. Overall, we show that information on floral biology can be automatically derived from image-based monitoring of flower plots.

@@

## Simulated data

To demonstrate our tracking algorithm and the effect of the three parameters, we used simulated data mimicking flower detections.To produce simulated detections of a flower, we defined a circle with radius rF and center point cF to delimit the boundaries of the flower. The first data point for the flower was then generated as a random point within this circle. Then, remaining data points (i.e. data points for the subsequent frames) were generated iteratively, each point constrained to be within the circle with radius rM and the center point of the previously generated point and within the flower circle.  This approach allowed us to simulate specific scenarios on which to demonstrate our algorithm by controlling the number of flowers generated and the number of data points produced per flower, the boundaries of a flower, and the maximum distance between current and next point (i.e. degree of movement between frames).

We generated four simulated data sets (Fig. 2) to demonstrate each of the three tracking parameters as well as the influence of frame rate on tracking reliability. In simulation A (Fig. 2A), we produced two flowers that are spatially distinct and temporally separated (i.e., one flower appears subsequent of the other). Each flower consists of ten data points and thus the data set spans over 20 frames. In simulation B, we produced four flowers that are spatially separated. Each flower consists of 5 data points, but the data set spans over 10 frames and each flower contains gaps of 1-3 data points at random temporal location. In simulation C, we produced four flowers that are in close proximity to each other and with a large degree of movement between frames. Each flower consist of 500 data points and the data sets spans over 500 frames. Finally, in simulation D, we produced four flowers consisting of a very high number of data points (10.000 per flower) but with only little movement between frames. Here, we used the same center point (cF) for all four flowers to increase overlap. Two pairs of flowers have a high degree of spatial overlap.

For each of the simulations, we present the tracking performance of our algorithm without the use of the three parameters and subsequently show how the parameters can be used to mitigate various challenges in the data and increase tracking accuracy. The two objects in simulation A are adjacent in time but spatially separated. We therefore introduce a limit to the spatial distance that is accepted between points in adjacent frames before they are considered unique objects. We do that with the max distance parameter. The objects in simulation B overlap in time but are well separated in space. However, each track contains gaps which can lead to track fragmentation. To allow gaps in track, we set a number of frames that a track can disappear before initiating a new track for points appearing in the same location. In simulation c, each object is allowed to move a lot between frames. Each object has a large spread of points and the four objects are in close proximity to each other. This means that between two frames, there is a substantial risk that a point for one object is closer to a point from another object rather than the point for that object, which will lead to identity shifts. To mitigate this, we calculate distances based on a current point and a set number of previous points in a track using the running mean parameter. This reduces the weighting on the displacement of points between the current and the previous frame. Finally, in simulation D, we have four objects that overlap in pairs of two, but the movement allowed between frames is very small, which means that a given point from an object is likely to be closest to the point from the same object in the previous frame. Effectively, while the overlap is substantial, the frame rate is sufficiently high that accurate tracking is ensured. To demonstrate the importance of frame rate for reliable tracking, we simulate a reduction in frame rate (i.e., increased degree of movement between frames) by basing tracking on every 10th, 20th, and 30th frame, respectively, and present these results.

![Figure 2: Simulated flower detection data for demonstration of our tracking algorithm and the effect of the three tracking parameters as well as the significance of frame rate for reliable tracking. A: 10 points per flower. rF = 200, rM = 100. B: 10 points per flower. rF = 200, rM = 100 C: 500 points per flower. rF = 200, rM = 200. D: 10,000 points per flower. rF = 100, rM = 1.](../figures/simulatedData.png){width="30%"}

## Common tracking errors and evaluating tracking performance

When tracking moving objects, two fundamental errors can occur. First, a single trajectory may be split into several tracks, an error known as fragmentation. This can happen when an object is temporarily occluded or have disappeared from view and a new track is erroneously initiated when it reappears, or when identity shifts occur between tracks, for example when they switch positions between frames. Fragmentation is especially prone in settings with high densities of objects. Second, merging occurs when several trajectories are assigned the same track identity. This can occur when a track is not correctly terminated at the end of a trajectory but rather continues on the trajectory of another object. Here, importantly, the two individual trajectories can be tracked without internal errors (i.e. without mismatches), but the error lies in the fact that the two trajectories have not been assigned unique tracking identities but instead are tracked as a single object.

@@

where fn, fp, and mm are the number of false negatives, false positives, and track mismatches, respectively, for frame t divided by the total number of ground truth annotations across all frames. The MOTA score ranges from -Inf to 1 with 1 being perfect detection and tracking. Tracking accuracy is commonly evaluated between manual annotations and detections produced, for example, by a convolutional neural network (CNN). Therefore, false positives and false negatives are included as errors in the MOTA score. In our case, we are not tracking on data produced by a CNN but rather on manually annotated tracks and do not have false positives or false negatives. Thus, these variables are ignored in the calculation of the MOTA score.

@@

## Flower data

### Study site and species

### The image series and flower annotations

Images were collected throughout the full growing seasons. While the original time-lapse frequency was higher, the basis for this study was 1-hour interval series. We limited these series to include only the flowering season and, to limit the workload for manual annotation, randomly sampled images from each series. The number of images for each series after sampling is given in Table 1. All flowers in bloom within each of these image were manually annotated using the rectangular bounding box tool in the VIA VGG annotation software [@dutta2019] and each individual flower was assigned a unique ID that was maintained throughout its lifetime. These annotations constitute our ground truth tracks. The series varied substantially in the number of flowers they contained (Table 1). Examples of annotated images from the 2019 NARS C series at three different times in the flowering season are given in figure 3. To demonstrate how individual flower tracking can result in detailed representations of floral phenology across the season, we use the ground truth annotations to show the variation in duration of flowering for each series.

# 
![Table 1: Number of images and number of individual flowers for each of the five time-lapse image series.](../figures/CameraOverview.png){width="40%"}

![Figure 3: Examples of annotated images for 2019 NARS C. A: Onset of flowering at DOY 145. B: Peak flowering at day of year (DOY) 151 (52 annotated flowers). C: 6 days after peak flowering (DOY 157). Flower bounding boxes are colour-coded by flower ID. D: Centroids of all flower annotations in the series, coloured by flower ID, showing how flowers may overlap in space.](../figures/AnnotationExamples_v2_4.png){width="20%"}

### Automatic flower tracking

We built a framework for tracking, filtering, and evaluating tracking of objects in time-lapse image series. The tracking algorithm (suppl. 1) is based on centroid tracking [@rahman2020; @bjerge2021]. That is, objects with minimum displacement between frames are associated. Centroids are derived from bounding boxes around objects stemming from manual annotation or automatic detection of objects. The algorithm has a set of user-adjusted parameters that can optimize tracking accuracy. The parameters are particularly relevant for tracking of objects that are constrained to a specific area such as flowers. However, the tracking algorithm could be used to track any objects. The speed of the tracking algorithm depends on the computational power available as well as the number of objects that are being tracked, but the time for computation will be negligible in most use-cases. Tracking of a series containing 85 objects ran at \~0.02 seconds per frame on a standard laptop. This, together with the fact that the algorithm handles frames sequentially, means that it can be applied both offline (on a set of detections/annotations that have already been produced) or online (real-time tracking frame per frame).

Several issues complicate the task of tracking individual flowers through time-lapse image series. First, as flowers move, they may temporarily occlude the view of other flowers and flowers close to the edge of the frame may move in and out of view between images. This can cause a track to be lost and fragmentation to occur (Fig. 4A). However, if a flower reappears in the same area as a flower is already being tracked after disappearing in only a few frames, it is a reasonable assumption that it is the same individual and not that the old flower wilted/disappeared and a new one developed. The parameter **max gap** sets the number of frames a flower can be absent before reappearing without the track ending and a new initiating. If a given flower has not been detected in a number of frames exceeding the max gap size, the track will be terminated. Concurrently, this can deal with potential false negatives, as a track that is missing a bounding box will not be lost.

Second, a standard centroid tracking algorithm will associate a point with a track based on minimum distance only, disregarding the absolute distance between centroids, which may lead to tracking errors (Fig. 4B). However, as flowers are constrained in their displacement, we can assume a maximum distance between points and force the initiation of tracks for points that exceed this threshold. We set this threshold with the parameter **max distance,** measured as the euclidean distance in number of pixels between objects in neighbouring frames.

Third, reliable tracking based on distance requires that movement of individuals are captured at sufficient frequency, such that each tracks are not switched between frames. Stand-alone image-based monitoring of the relatively slow process of flower development through full growing seasons (several months) is done with low frame rate, limited by power consumption and storage capacity both on and off the camera. Here, we used several images per day and while this temporal resolution captures the phenology of the flowers at very high detail, they fail to capture the full movement of flowers. For example, as the wind shifts, the flower heads change direction, which in the image series happens instantaneously (i.e., between two consecutive frames) (Fig. 4C). The fact that the frame rate is insufficient to capture smooth movement means that bounding boxes around a single individual in two consecutive frames cannot be associated based on overlap (e.g. intersection over union) as there often is no overlap. With a high number of flowers in close vicinity to each other, there's a substantial risk of identity switching between flowers as they move among each other (Fig. 1C;4C). Additionally, as the flowers are constrained by their stalk, there is a limit to the distance they can move, which prevents prediction of future location of an individual based on its previous trajectory. Here, we introduce the tracking parameter **running mean**, with which we can base tracking on the distance between a point in the current frame and the running mean of the positions of a number of previous points in a track, taking advantage of the fact that a flower moves around a central point. Increasing the value for this parameter decreases the weighting of movement between individual frames. Setting the parameter to 1 means that tracking is based on the distance to the centroid in the previous frame only.

![Figure 4: Centroid tracking may produce erroneous associations when objects move or disappear between frames. A: A track may be lost when a flower temporarily disappears from view. If gaps in tracks are not allowed during tracking, a new track will erroneously be initiated when the flower reappears. B: A flower may be linked to an incorrect track despite large distance between them. Here, a flower disappears from view and another reappears between two frames. If no threshold for maximum distance between an object and a track is set, the reappearing flower will be linked to the track of the disappearing flower based on shortest distance despite the absolute distance being large. C: A flower may be linked to and incorrect track despite shortest distance to the running mean of the correct track. As flowers move around a central point, it is relevant to base tracking on a running mean of previous positions. Dots show position of flower in previous frame. Crosses show running mean of positions in x \> 1 number of previous frames. Dots and crosses coloured according to track identity.](../figures/errorFigure.png){width="100%"}

#### Input data for the tracking algorithm

#### Identifying optimal user parameters

To explore the effect of the three tracking parameters and to identify the optimal combination of parameters for our use-case of tracking flowers, we applied the tracking algorithm on each of the five image series with every combination of a range of values for each parameter (3179 combinations): 0-160 with a step size of 10 for max gap; 1-160 with a step size of 10 for running mean; 0-1000 with a step size of 100 for max distance. Note that max distance set to zero ignores the parameter altogether. We identified the setting(s) that returned the most accurate tracks for each series and compare the tracking results between optimal settings and all parameters ignored (max gap = 0, running mean = 1, max distance = 0).

### Identifying the most reliably tracked flowers

While the implemented tracking algorithm produced highly accurate flower tracks, it did not do so without errors. Many flowers could be tracked with certainty or near-certainty, but some settings proved challenging, especially series with many flowers or flowers in close proximity to each other. In order to be able to identify the most reliably tracked flowers in a series without the need for manual annotation, we use a conservative filtering method that extracts the most trustworthy tracks from a scene based on spatial distance between tracks. We derive the distance between tracks based on the centroids of points included in a track produced by the tracking algorithm. When tracks consist of only a single point, the coordinates of the point is used as the centroid. For tracks consisting of two points we use the midpoint of the line between them. For tracks consisting of three points or more, we calculate the convex hull of all the points included in the track and use the centroid of this polygon. For the special case where a track contains more than two points but where the points are collinear (i.e., lie on a single straight line), we use the midpoint of the line through the points. We hereafter refer to these centroids as track centroids.

@@

## Simulated data

Tracking for simulation A without applying the three tracking parameters resulted in a merging error (i.e. the two objects were tracked as one). Setting the max distance to 200 removed this error and resulted in perfect tracking of the two objects. Tracking for simulation B without applying the tracking parameters returned 8 tracks and 9 mismatches (MOTA = 0.55). Setting max gap to 4 decreased the number of mismatches to 4 (4 tracks, MOTA = 0.8) and including a value of 200 for max distance as for simulation A removes the remaining errors and resulted in perfect tracking. For simulation C, tracking without parameters ignored returned the correct number of tracks (4), but 38 mismatches were made (MOTA = 0.98). Setting the running mean parameter to 20, 40, 60, resulted in 16, 8, and 4 mismatches, respectively, while a running mean value of 80 resulted in perfect tracking. A running mean value of 100 resulted in 4 mismatches. For simulation D, the algorithm tracked the four objects perfectly despite substantial overlap. Basing tracking on every 10th, 20th, and 30th frame resulted in 2, 4, and 10 mismatches, respectively.
